### 2017-11-14 submission/manuscript.Rmd
add Will's abstract suggestions

**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer. Important drivers within this axis have ranged from individual microbes to the whole community. A recent meta-analysis investigated whether consistent biomarkers could be identified across studies. This report expands on this previous research and tests the hypothesis that the bacterial community is an important driver of disease from control to adenoma to carcinoma. To test this hypothesis we examined both feces (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) across 14 different studies.

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) while only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk. Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification AUC for both adenoma and carcinoma for fecal and tissue samples. For the studies that were analyzed, most were adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma microbiota research.

**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis. The evidence that has been collated within this study is much stronger for carcinoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

>>>>

**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer (CRC). Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r sum(filter(tissue_n, study == "flemer")[, c("control", "adenoma", "cancer")]) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. For the included studies, most were adequately powered for large effect size differences. This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.

**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC pathogenesis is much stronger for carcinoma then adenoma. A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

### 2017-11-14 submission/manuscript.tex
add Will's abstract suggestions

\textbf{Background.} An increasing body of literature suggests that
there is a role for the microbiota in colorectal cancer. Important
drivers within this axis have ranged from individual microbes to the
whole community. A recent meta-analysis investigated whether consistent
biomarkers could be identified across studies. This report expands on
this previous research and tests the hypothesis that the bacterial
community is an important driver of disease from control to adenoma to
carcinoma. To test this hypothesis we examined both feces (total
individuals = 1737) and tissue (total samples = 492) across 14 different
studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Lower Shannon diversity and evenness in fecal samples
resulted in a significant increase in relative risk for carcinoma
(P-value \textless{} 0.05) while only evenness for adenoma (P-value
\textless{} 0.05) resulted in a slightly increased relative risk.
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern with a significantly increased relative risk
by their presence for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value \textless{} 0.05). Using the whole
community versus only CRC associated genera to build a prediction model
resulted in a higher classification AUC for both adenoma and carcinoma
for fecal and tissue samples. For the studies that were analyzed, most
were adequately powered for large effect size differences which may be
more amenable for carcinoma than for adenoma microbiota research.

\textbf{Conclusions.} This data provides support for the importance of
the bacteral community to both adenoma and carcinoma genesis. The
evidence that has been collated within this study is much stronger for
carcinoma and this may be in part due to the low power to detect more
subtle changes in the majority of studies that have been performed to
date.

>>>>

\textbf{Background.} An increasing body of literature suggests that
there is a role for the microbiota in colorectal cancer (CRC). Important
drivers within this context have ranged from individual microbes to the
whole community. Our study expands on a recent meta-analysis
investigating microbial biomarkers for CRC by testing the hypothesis
that the bacterial community is an important driver of both early
(adenoma) and late (carcinoma) stage of disease. To test this hypothesis
we examined both feces (n = 1737) and tissue (492 total samples from 400
individuals) across 14 different studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
in a slightly increased relative risk while lower Shannon diversity and
evenness in fecal samples resulted in a significant increase in relative
risk for carcinoma (P-value \textless{} 0.05). Previously associated
colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
pattern with a significantly increased relative risk by their presence
for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
\textless{} 0.05). Using the whole community versus only CRC associated
genera to build a prediction model resulted in a higher classification
based on Area Under the Curve (AUC) for both adenoma and carcinoma using
fecal and tissue samples. For the included studies, most were adequately
powered for large effect size differences. This may be more amenable for
carcinoma than for adenoma microbiota research due to the smaller
community level changes observed in our results.

\textbf{Conclusions.} This data provides support for the importance of
the bacteral community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis is much stronger for carcinoma then adenoma. A strong
reason for this may be in part due to the low power to detect more
subtle changes in the majority of studies that have been performed to
date.
